---
title: "Following Best Practices for Using eBird Data Tutorial"
output: 
  pdf_document:
    latex_engine: xelatex
---

When looking for machine and statistical learning project, I ran across [Best Practices for Using eBird Data](https://cornelllabofornithology.github.io/ebird-best-practices/index.html). It walks through using random forest to predict bird encounter rates for a given species in a given region.  To properly understand this, I'm walking through the tutorial.

<<<<<<< HEAD
<<<<<<< HEAD
# Introduction

=======
>>>>>>> 8ddaf72676f352482826582b05ee6c0209c69ffa
=======
>>>>>>> 8ddaf72676f352482826582b05ee6c0209c69ffa
```{r}
install.packages("remotes")
remotes::install_github("mstrimas/ebppackages")
```

This didn't work, so I assume that we'll be using _rebird_ and _auk_ and import the packages as we come acorss them.

<<<<<<< HEAD
<<<<<<< HEAD
```{r}
library(dplyr)
library(auk)
library(sf)
library(rnaturalearth)
library(tidyselect)
```

This recommends getting the full ebird text file, but that is very large.  We instead requested data on Northern Cardinal sightings in Illinois.

Set the path to our data
```{r}
# set ebd path
auk_set_ebd_path("../data/raw/ebd")
```

I got errors while trying to install 'sf', and used the following to resolve it:
* https://community.rstudio.com/t/configuration-failed-for-package-units/76417
  * sudo apt install libudunits2-dev
* https://mothergeo-py.readthedocs.io/en/latest/development/how-to/gdal-ubuntu-pkg.html
* https://stackoverflow.com/questions/44973639/trouble-installing-sf-due-to-gdal 
  * gdalinfo --version
  * sudo apt install gdalinfo

Replace bird conservation area code 27 with 22, which includes Chicago.
```{r}
# file to save spatial data
gpkg_dir <- "../data/raw"
if (!dir.exists(gpkg_dir))(
  dir.create(gpkg_dir)
)

f_ne <- file.path(gpkg_dir,"gis-data.gpkg")

#download bcrs
# tmp_dir <- normalizePath(tempdir())
# tmp_bcr <- file.path(tmp_dir, "bcr.zip")
# paste0("https://www.birdscanada.org/research/gislab/download/", 
#        "bcr_terrestrial_shape.zip") %>% 
#   download.file(destfile = tmp_bcr)
# unzip(tmp_bcr, exdir = tmp_dir)
bcr <- file.path("../data/raw", "BCR_Terrestrial_master_International.shp") %>%
  read_sf() %>%
  select(bcr_code = BCR, bcr_name = LABEL) %>%
  dplyr::filter(bcr_code == 22)

```

The Birds Canada links have changed, so I went to [the site](https://www.birdscanada.org/bird-science/nabci-bird-conservation-regions/) and downlaoded the terrestrial shape file directly instead.

# eBird Data

Switching in the name of the file I have
```{r}
library(auk)
library(lubridate)
library(sf)
library(gridExtra)
library(tidyverse)
# resolve namespace conflicts
select <- dplyr::select

# setup data directory
dir.create("../data/raw/ebd", showWarnings = FALSE)

ebd <- auk_ebd("ebd_US_norcar_relOct-2020.txt"
               ,
               file_sampling = "ebd_sampling_relOct-2020.txt"
               )

# ebd <- auk_ebd("ebd_US_norcar_relOct-2020.txt")

```

```{r}

# political boundaries
# land border with lakes removed
ne_land <- ne_download(scale = 50, category = "cultural",
                       type = "admin_0_countries_lakes",
                       returnclass = "sf") %>%
  filter(CONTINENT == "North America") %>%
  st_set_precision(1e6) %>%
  st_union()
# country lines
# downloaded globally then filtered to north america with st_intersect()
ne_country_lines <- ne_download(scale = 50, category = "cultural",
                                type = "admin_0_boundary_lines_land",
                                returnclass = "sf") %>% 
  st_geometry()
ne_country_lines <- st_intersects(ne_country_lines, ne_land, sparse = FALSE) %>%
  as.logical() %>%
  {ne_country_lines[.]}
# states, north america
ne_state_lines <- ne_download(scale = 50, category = "cultural",
                              type = "admin_1_states_provinces_lines",
                              returnclass = "sf") %>%
  filter(adm0_a3 %in% c("USA", "CAN")) %>%
  mutate(iso_a2 = recode(adm0_a3, USA = "US", CAN = "CAN")) %>% 
  select(country = adm0_name, country_code = iso_a2)


# output
unlink(f_ne)
write_sf(ne_land, f_ne, "ne_land")
write_sf(ne_country_lines, f_ne, "ne_country_lines")
write_sf(ne_state_lines, f_ne, "ne_state_lines")
write_sf(bcr, f_ne, "bcr")
```

The map of bird conservation areas can be seen [here](https://nabci-us.org/resources/bird-conservation-regions-map/).
```{r}
ebd_filters <- ebd %>% 
  auk_species("Northern cardinal") %>% 
  # Eastern Tallgrass Prairie bcr
  auk_bcr(bcr = 22) %>% 
  # june, use * to get data from any year
  auk_date(date = c("*-06-01", "*-06-30")) %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()
ebd_filters

```

```{r}
# output files
data_dir <- "../data/processed"
if (!dir.exists(data_dir)) {
  dir.create(data_dir)
}
f_ebd <- file.path(data_dir, "ebd_cardinal_june_bcr22.txt")
f_sampling <- file.path(data_dir, "ebd_checklists_june_bcr22.txt")

# only run if the files don't already exist
if (!file.exists(f_ebd)) {
  auk_filter(ebd_filters, file = f_ebd, file_sampling = f_sampling)
}

# if (!file.exists(f_ebd)) {
#   auk_filter(ebd_filters, file = f_ebd)
# }

```

```{r}
ebd_zf <- auk_zerofill(f_ebd, f_sampling, collapse = TRUE)
# ebd_zf <- read_ebd(f_ebd)
ebd_zf
```

```{r}
# function to convert time observation to hours since midnight
time_to_decimal <- function(x) {
  x <- hms(x, quiet = TRUE)
  hour(x) + minute(x) / 60 + second(x) / 3600
}

# clean up variables
ebd_zf <- ebd_zf %>% 
  mutate(
    # convert X to NA
    observation_count = if_else(observation_count == "X", 
                                NA_character_, observation_count),
    observation_count = as.integer(observation_count),
    # effort_distance_km to 0 for non-travelling counts
    effort_distance_km = if_else(protocol_type != "Traveling", 
                                 0, effort_distance_km),
    # convert time to decimal hours since midnight
    time_observations_started = time_to_decimal(time_observations_started),
    # split date into year and day of year
    year = year(observation_date),
    day_of_year = yday(observation_date)
  )

```

## 2.4 Accounting for variation in detectability

```{r}
ebd_zf_filtered <- ebd_zf %>% 
  filter(
    # effort filters
    duration_minutes <= 5 * 60,
    effort_distance_km <= 5,
    # last 10 years of data
    year >= 2010,
    # 10 or fewer observers
    number_observers <= 10)

# ebd_zf_filtered <- ebd_zf_filtered %>%
#   mutate(species_observed = case_when(observation_count > 1 ~ TRUE, TRUE ~ FALSE))
```

```{r}
ebird <- ebd_zf_filtered %>% 
  select(checklist_id, observer_id, sampling_event_identifier,
         scientific_name,
         observation_count, species_observed, 
         state_code, locality_id, latitude, longitude,
         protocol_type, all_species_reported,
         observation_date, year, day_of_year,
         time_observations_started, 
         duration_minutes, effort_distance_km,
         number_observers)
write_csv(ebird, "../data/processed/ebd_cardinal_june_bcr22_zf.csv", na = "")
```

## 2.5 Exploratory analysis and visualization

```{r}
head(ebird)

# load and project gis data
new_vector = st_read(f_ne)

map_proj <- st_crs(new_vector)
ne_land <- read_sf(f_ne, "ne_land") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
bcr <- read_sf(f_ne, "bcr") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
ne_country_lines <- read_sf(f_ne, "ne_country_lines") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
ne_state_lines <- read_sf(f_ne, "ne_state_lines") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()

# prepare ebird data for mapping
ebird_sf <- ebird %>% 
  # convert to spatial points
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(crs = map_proj) %>% 
  select(species_observed)

# map
par(mar = c(0.25, 0.25, 0.25, 0.25))
# set up plot area
plot(st_geometry(ebird_sf), col = NA)
# contextual gis data
plot(ne_land, col = "#dddddd", border = "#888888", lwd = 0.5, add = TRUE)
plot(bcr, col = "#cccccc", border = NA, add = TRUE)
plot(ne_state_lines, col = "#ffffff", lwd = 0.75, add = TRUE)
plot(ne_country_lines, col = "#ffffff", lwd = 1.5, add = TRUE)
# ebird observations
# not observed
plot(st_geometry(ebird_sf),
     pch = 19, cex = 0.1, col = alpha("#555555", 0.25),
     add = TRUE)
# observed
plot(filter(ebird_sf, species_observed) %>% st_geometry(),
     pch = 19, cex = 0.3, col = alpha("#4daf4a", 1),
     add = TRUE)
# legend
legend("bottomright", bty = "n",
       col = c("#555555", "#4daf4a"),
       legend = c("eBird checklists", "Northern Cardinal sightings"),
       pch = 19)
box()
par(new = TRUE, mar = c(0, 0, 3, 0))
title("Northern eBird Observations\nJune 2010-2020, BCR 22")
```

```{r}
library(sf)
library(raster)
library(MODIS)
library(exactextractr)
library(viridis)
library(tidyverse)
# resolve namespace conflicts
select <- dplyr::select
map <- purrr::map
projection <- raster::projection

# bcr 22 boundary
bcr <- read_sf("../data/raw/gis-data.gpkg", "bcr") %>% 
  filter(bcr_code == 22) %>% 
  # project to the native modis projection
  st_transform(crs = paste("+proj=sinu +lon_0=0 +x_0=0 +y_0=0",
                           "+a=6371007.181 +b=6371007.181 +units=m +no_defs"))
# load ebird data
ebird <- read_csv("../data/processed/ebd_cardinal_june_bcr22_zf.csv")
# get list of tiles required to cover this bcr
tiles <- getTile(bcr)
tiles@tile
#> [1] "h10v06" "h10v05" "h11v05"
```


```{r landcover-dl-r, eval = FALSE}
# earliest year of ebird data
begin_year <- format(min(ebird$observation_date), "%Y.01.01")
# end date for ebird data
end_year <- format(max(ebird$observation_date), "%Y.12.31")
# download tiles and combine into a single raster for each year
tifs <- runGdal(product = "MCD12Q1", collection = "006", SDSstring = "01", 
                extent = bcr %>% st_buffer(dist = 10000), 
                begin = begin_year, end = end_year, 
                outDirPath = "../data/raw", job = "modis",
                MODISserverOrder = "LPDAAC") %>%
  pluck("MCD12Q1.006") %>% 
  unlist()
# rename tifs to have more descriptive names
new_names <- format(as.Date(names(tifs)), "%Y") %>% 
  sprintf("modis_mcd12q1_umd_%s.tif", .) %>% 
  file.path(dirname(tifs), .)
file.rename(tifs, new_names)
```
```{r}
# load the landcover data
landcover <- list.files("../data/raw/modis", "^modis_mcd12q1_umd", 
                        full.names = TRUE) %>% 
  stack()
# label layers with year
landcover <- names(landcover) %>% 
  str_extract("(?<=modis_mcd12q1_umd_)[0-9]{4}") %>% 
  paste0("y", .) %>% 
  setNames(landcover, .)
landcover
#> class      : RasterStack 
#> dimensions : 1969, 4301, 8468669, 9  (nrow, ncol, ncell, nlayers)
#> resolution : 463, 463  (x, y)
#> extent     : -8745491, -6752783, 3242262, 4154525  (xmin, xmax, ymin, ymax)
#> crs        : +proj=sinu +lon_0=0 +x_0=0 +y_0=0 +a=6371007.181 +b=6371007.181 +units=m +no_defs 
#> names      : y2010, y2011, y2012, y2013, y2014, y2015, y2016, y2017, y2018 
#> min values :     0,     0,     0,     0,     0,     0,     0,     0,     0 
#> max values :   255,   255,   255,   255,   255,   255,   255,   255,   255
```

```{r}
max_lc_year <- names(landcover) %>% 
  str_extract("[0-9]{4}") %>% 
  as.integer() %>% 
  max()
```

```{r}
neighborhood_radius <- 5 * ceiling(max(res(landcover))) / 2
ebird_buff <- ebird %>% 
  distinct(year = format(observation_date, "%Y"),
           locality_id, latitude, longitude) %>% 
  # for 2019 use 2018 landcover data
  mutate(year_lc = if_else(as.integer(year) > max_lc_year, 
                           as.character(max_lc_year), year),
         year_lc = paste0("y", year_lc)) %>% 
  # convert to spatial features
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  # transform to modis projection
  st_transform(crs = projection(landcover)) %>% 
  # buffer to create neighborhood around each point
  st_buffer(dist = neighborhood_radius) %>% 
  # nest by year
  nest(data = c(year, locality_id, geometry))
```

```{r}
# function to summarize landcover data for all checklists in a given year
calculate_pland <- function(yr, regions, lc) {
  locs <- st_set_geometry(regions, NULL)
  exact_extract(lc[[yr]], regions, progress = FALSE) %>% 
    map(~ count(., landcover = value)) %>% 
    tibble(locs, data = .) %>% 
    unnest(data)
}
# iterate over all years extracting landcover for all checklists in each
lc_extract <- ebird_buff %>% 
  mutate(pland = map2(year_lc, data, calculate_pland, lc = landcover)) %>% 
  select(pland) %>% 
  unnest(cols = pland)
```

```{r}
pland <- lc_extract %>% 
  # calculate proporiton
  group_by(locality_id, year) %>% 
  mutate(pland = n / sum(n)) %>% 
  ungroup() %>% 
  select(-n) %>% 
  # remove NAs after tallying so pland is relative to total number of cells
  filter(!is.na(landcover))
```

```{r}
# convert names to be more descriptive
lc_names <- tibble(landcover = 0:15,
                   lc_name = c("pland_00_water", 
                               "pland_01_evergreen_needleleaf", 
                               "pland_02_evergreen_broadleaf", 
                               "pland_03_deciduous_needleleaf", 
                               "pland_04_deciduous_broadleaf", 
                               "pland_05_mixed_forest",
                               "pland_06_closed_shrubland", 
                               "pland_07_open_shrubland", 
                               "pland_08_woody_savanna", 
                               "pland_09_savanna", 
                               "pland_10_grassland", 
                               "pland_11_wetland", 
                               "pland_12_cropland", 
                               "pland_13_urban", 
                               "pland_14_mosiac", 
                               "pland_15_barren"))
pland <- pland %>% 
  inner_join(lc_names, by = "landcover") %>% 
  arrange(landcover) %>% 
  select(-landcover)

# tranform to wide format, filling in implicit missing values with 0s%>% 
pland <- pland %>% 
  pivot_wider(names_from = lc_name, 
              values_from = pland, 
              values_fill = list(pland = 0))

# save
write_csv(pland, "../data/processed/modis_pland_location-year.csv")
```

```{r}
agg_factor <- round(2 * neighborhood_radius / res(landcover))
r <- raster(landcover) %>% 
  aggregate(agg_factor) 
r <- bcr %>% 
  st_transform(crs = projection(r)) %>% 
  rasterize(r, field = 1) %>% 
  # remove any empty cells at edges
  trim()
r <- writeRaster(r, filename = "../data/processed/prediction-surface.tif", overwrite = TRUE)

```


```{r}
# get cell centers and create neighborhoods
r_centers <- rasterToPoints(r, spatial = TRUE) %>% 
  st_as_sf() %>% 
  transmute(id = row_number())
r_cells <- st_buffer(r_centers, dist = neighborhood_radius)

# extract landcover values within neighborhoods, only needed most recent year
lc_extract_pred <- landcover[[paste0("y", max_lc_year)]] %>% 
  exact_extract(r_cells, progress = FALSE) %>% 
  map(~ count(., landcover = value)) %>% 
  tibble(id = r_cells$id, data = .) %>% 
  unnest(data)

# calculate the percent for each landcover class
pland_pred <- lc_extract_pred %>% 
  count(id, landcover) %>% 
  group_by(id) %>% 
  mutate(pland = n / sum(n)) %>% 
  ungroup() %>% 
  select(-n) %>% 
  # remove NAs after tallying so pland is relative to total number of cells
  filter(!is.na(landcover))

# convert names to be more descriptive
pland_pred <- pland_pred %>% 
  inner_join(lc_names, by = "landcover") %>% 
  arrange(landcover) %>% 
  select(-landcover)

# tranform to wide format, filling in implicit missing values with 0s
pland_pred <- pland_pred %>% 
  pivot_wider(names_from = lc_name, 
              values_from = pland, 
              values_fill = list(pland = 0)) %>% 
  mutate(year = max_lc_year) %>% 
  select(id, year, everything())

# join in coordinates
pland_coords <- st_transform(r_centers, crs = 4326) %>% 
  st_coordinates() %>% 
  as.data.frame() %>% 
  cbind(id = r_centers$id, .) %>% 
  rename(longitude = X, latitude = Y) %>% 
  inner_join(pland_pred, by = "id")
```

```{r}
forest_cover <- pland_coords %>% 
  # convert to spatial features
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(crs = projection(r)) %>% 
  # rasterize points
  rasterize(r, field = "pland_04_deciduous_broadleaf") %>% 
  # project to albers equal-area for mapping
  projectRaster(crs = st_crs(102003)$proj4string, method = "ngb") %>% 
  # trim off empty edges of raster
  trim()

# make a map
par(mar = c(0.25, 0.25, 2, 0.25))
t <- str_glue("Proportion of Deciduous Broadleaf Forest\n",
              "{max_lc_year} MODIS Landcover")
plot(forest_cover, axes = FALSE, box = FALSE, col = viridis(10), main = t)
```

```{r}
elev <- raster("../data/raw/elevation_1KMmd_GMTEDmd.tif")
# crop, buffer bcr by 10 km to provide a little wiggly room
elev <- bcr %>% 
  st_buffer(dist = 10000) %>% 
  st_transform(crs = projection(elev)) %>% 
  crop(elev, .) %>% 
  projectRaster(crs = projection(landcover))
```

```{r}
# buffer each checklist location
ebird_buff_noyear <- ebird %>% 
  distinct(locality_id, latitude, longitude) %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(crs = projection(elev)) %>% 
  st_buffer(dist = neighborhood_radius)

# extract elevation values and calculate median and sd
locs <- st_set_geometry(ebird_buff_noyear, NULL) %>% 
    mutate(id = row_number())
elev_checklists <- exact_extract(elev, ebird_buff_noyear, progress = FALSE) %>% 
  map_dfr(~ tibble(elevation_median = mean(.$value, na.rm = TRUE),
                   elevation_sd = sd(.$value, na.rm = TRUE))) %>% 
  # join to lookup table to get locality_id
  bind_cols(locs, .)
```

```{r}
# extract and calculate median and sd
elev_pred <- exact_extract(elev, r_cells, progress = FALSE) %>% 
  map_dfr(~ tibble(elevation_median = mean(.$value, na.rm = TRUE),
                   elevation_sd = sd(.$value, na.rm = TRUE))) %>% 
  # join to lookup table to get locality_id
  bind_cols(st_drop_geometry(r_cells), .)
```

```{r}
# extract and calculate median and sd
elev_pred <- exact_extract(elev, r_cells, progress = FALSE) %>% 
  map_dfr(~ tibble(elevation_median = mean(.$value, na.rm = TRUE),
                   elevation_sd = sd(.$value, na.rm = TRUE))) %>% 
  # join to lookup table to get locality_id
  bind_cols(st_drop_geometry(r_cells), .)
```

```{r}
# checklist covariates
pland_elev_checklist <- inner_join(pland, elev_checklists, by = "locality_id")
write_csv(pland_elev_checklist, "../data/processed/pland-elev_location-year.csv")

# prediction surface covariates
pland_elev_pred <- inner_join(pland_coords, elev_pred, by = "id")
write_csv(pland_elev_pred, "../data/processed/pland-elev_prediction-surface.csv")
glimpse(pland_elev_pred)
#> Rows: 90,949
#> Columns: 22
#> $ id                            <int> 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25…
#> $ longitude                     <dbl> -77.3, -77.3, -77.3, -77.3, -77.3, -77.2, -77.4, -77.3, -77.3, -77.3, -77.3, -77.2, -77.2…
#> $ latitude                      <dbl> 37.2, 37.2, 37.2, 37.2, 37.2, 37.2, 37.2, 37.2, 37.2, 37.2, 37.2, 37.2, 37.2, 37.2, 37.2,…
#> $ year                          <int> 2018, 2018, 2018, 2018, 2018, 2018, 2018, 2018, 2018, 2018, 2018, 2018, 2018, 2018, 2018,…
#> $ pland_00_water                <dbl> 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,…
#> $ pland_01_evergreen_needleleaf <dbl> 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0…
#> $ pland_02_evergreen_broadleaf  <dbl> 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,…
#> $ pland_03_deciduous_needleleaf <dbl> 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,…
#> $ pland_04_deciduous_broadleaf  <dbl> 0.0000, 0.1724, 0.0345, 0.0000, 0.0000, 0.0345, 0.0000, 0.0000, 0.0000, 0.0000, 0.0345, 0…
#> $ pland_05_mixed_forest         <dbl> 0.0000, 0.0000, 0.0000, 0.1379, 0.1379, 0.2759, 0.0000, 0.0000, 0.0345, 0.1379, 0.0345, 0…
#> $ pland_06_closed_shrubland     <dbl> 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,…
#> $ pland_07_open_shrubland       <dbl> 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,…
#> $ pland_08_woody_savanna        <dbl> 0.966, 0.828, 0.724, 0.690, 0.862, 0.690, 0.621, 0.966, 0.793, 0.828, 0.793, 0.310, 0.552…
#> $ pland_09_savanna              <dbl> 0.0000, 0.0000, 0.2069, 0.1724, 0.0000, 0.0000, 0.0345, 0.0345, 0.1379, 0.0345, 0.1379, 0…
#> $ pland_10_grassland            <dbl> 0.0345, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0345, 0.0000, 0.0000, 0…
#> $ pland_11_wetland              <dbl> 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,…
#> $ pland_12_cropland             <dbl> 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0.0000, 0…
#> $ pland_13_urban                <dbl> 0.0000, 0.0000, 0.0345, 0.0000, 0.0000, 0.0000, 0.3448, 0.0000, 0.0000, 0.0000, 0.0000, 0…
#> $ pland_14_mosiac               <dbl> 0.000, 0.000, 0.000, 0.000, 0.000, 0.000, 0.000, 0.000, 0.000, 0.000, 0.000, 0.000, 0.000…
#> $ pland_15_barren               <dbl> 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,…
#> $ elevation_median              <dbl> 40.0, 41.8, 42.6, 42.8, 42.5, 42.8, 45.6, 44.7, 43.6, 41.2, 40.6, 40.9, 42.8, 78.0, 68.1,…
#> $ elevation_sd                  <dbl> 3.75, 2.78, 2.41, 3.28, 1.40, 1.66, 2.74, 3.33, 2.29, 2.32, 2.52, 3.41, 1.53, 8.61, 7.02,…
```
=======
This recommends getting the full ebird text file, but 
>>>>>>> 8ddaf72676f352482826582b05ee6c0209c69ffa
=======
This recommends getting the full ebird text file, but 
>>>>>>> 8ddaf72676f352482826582b05ee6c0209c69ffa

# 4
## 4.2
```{r}
library(sf)
library(raster)
library(dggridR)
library(lubridate)
library(ranger)
library(scam)
library(PresenceAbsence)
library(verification)
library(ebirdst)
library(fields)
library(gridExtra)
library(tidyverse)
# resolve namespace conflicts
select <- dplyr::select
map <- purrr::map
projection <- raster::projection

# set random number seed to insure fully repeatable results
set.seed(1)

# setup output directory for saved results
if (!dir.exists("../output")) {
  dir.create("../output")
}

# ebird data
ebird <- read_csv("../data/processed/ebd_cardinal_june_bcr22_zf.csv") %>% 
  # year required to join to habitat data
  mutate(year = year(observation_date))

# modis habitat covariates
habitat <- read_csv("../data/processed/pland-elev_location-year.csv") %>% 
  mutate(year = as.integer(year))

# combine ebird and habitat data
ebird_habitat <- inner_join(ebird, habitat, by = c("locality_id", "year"))

# prediction surface
pred_surface <- read_csv("../data/processed/pland-elev_prediction-surface.csv")
# latest year of landcover data
max_lc_year <- max(pred_surface$year)
r <- raster("../data/processed/prediction-surface.tif")

# load gis data for making maps
map_proj <- st_crs(102003)
ne_land <- read_sf("../data/raw/gis-data.gpkg", "ne_land") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
bcr <- read_sf("../data/raw/gis-data.gpkg", "bcr") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
ne_country_lines <- read_sf("../data/raw/gis-data.gpkg", "ne_country_lines") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
ne_state_lines <- read_sf("../data/raw/gis-data.gpkg", "ne_state_lines") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
```

## 4.3
```{r}
# bounding box to generate points from
bb <- st_bbox(c(xmin = -0.1, xmax = 0.1, ymin = -0.1, ymax = 0.1), 
              crs = 4326) %>% 
  st_as_sfc() %>% 
  st_sf()
# random points
pts <- st_sample(bb, 500) %>% 
  st_sf(as.data.frame(st_coordinates(.)), geometry = .) %>% 
  rename(lat = Y, lon = X)

# contruct a hexagonal grid with ~ 5 km between cells
dggs <- dgconstruct(spacing = 5)
# for each point, get the grid cell
pts$cell <- dgGEO_to_SEQNUM(dggs, pts$lon, pts$lat)$seqnum

# sample one checklist per grid cell
pts_ss <- pts %>% 
  group_by(cell) %>% 
  sample_n(size = 1) %>% 
  ungroup()

# generate polygons for the grid cells
hexagons <- dgcellstogrid(dggs, unique(pts$cell), frame = FALSE) %>% 
  st_as_sf()
ggplot() +
  geom_sf(data = hexagons) +
  geom_sf(data = pts, size = 0.5) +
  geom_sf(data = pts_ss, col = "red") +
  theme_bw()
```

```{r}
# generate hexagonal grid with ~ 5 km betweeen cells
dggs <- dgconstruct(spacing = 5)
# get hexagonal cell id and week number for each checklist
checklist_cell <- ebird_habitat %>% 
  mutate(cell = dgGEO_to_SEQNUM(dggs, longitude, latitude)$seqnum,
         year = year(observation_date),
         week = week(observation_date))
# sample one checklist per grid cell per week
# sample detection/non-detection independently 
ebird_ss <- checklist_cell %>% 
  group_by(species_observed, year, week, cell) %>% 
  sample_n(size = 1) %>% 
  ungroup()
```

```{r}
# original data
nrow(ebird_habitat)
names(ebird_habitat)
#> [1] 125202
count(ebird_habitat, species_observed) %>% 
  mutate(percent = n / sum(n))
#> # A tibble: 2 x 3
#>   species_observed     n percent
#>   <lgl>            <int>   <dbl>
#> 1 FALSE            46355  0.957 
#> 2 TRUE              2095  0.0432

# after sampling
nrow(ebird_ss)
names(ebird_ss)
head(ebird_ss)
#> [1] 20093
count(ebird_ss, species_observed) %>% 
  mutate(percent = n / sum(n))
#> # A tibble: 2 x 3
#>   species_observed     n percent
#>   <lgl>            <int>   <dbl>
#> 1 FALSE            18617  0.927 
#> 2 TRUE              1476  0.0735
```

```{r}
# convert checklists to spatial features
all_pts <- ebird_habitat %>%  
  st_as_sf(coords = c("longitude","latitude"), crs = 4326) %>%
  st_transform(crs = map_proj) %>% 
  select(species_observed)
ss_pts <- ebird_ss %>%  
  st_as_sf(coords = c("longitude","latitude"), crs = 4326) %>%
  st_transform(crs = map_proj) %>% 
  select(species_observed)
both_pts <- list(before_ss = all_pts, after_ss = ss_pts)

```

```{r}
# map
p <- par(mfrow = c(2, 1))
for (i in seq_along(both_pts)) {
  par(mar = c(0.25, 0.25, 0.25, 0.25))
  # set up plot area
  plot(st_geometry(both_pts[[i]]), col = NA)
  # contextual gis data
  plot(ne_land, col = "#dddddd", border = "#888888", lwd = 0.5, add = TRUE)
  plot(bcr, col = "#cccccc", border = NA, add = TRUE)
  plot(ne_state_lines, col = "#ffffff", lwd = 0.75, add = TRUE)
  plot(ne_country_lines, col = "#ffffff", lwd = 1.5, add = TRUE)
  # ebird observations
  # not observed
  plot(st_geometry(both_pts[[i]]),
       pch = 19, cex = 0.1, col = alpha("#555555", 0.25),
       add = TRUE)
  # observed
  plot(filter(both_pts[[i]], species_observed) %>% st_geometry(),
       pch = 19, cex = 0.3, col = alpha("#4daf4a", 0.5),
       add = TRUE)
  # legend
  legend("bottomright", bty = "n",
         col = c("#555555", "#4daf4a"),
         legend = c("Non-detection", "Detection"),
         pch = 19)
  box()
  par(new = TRUE, mar = c(0, 0, 3, 0))
  if (names(both_pts)[i] == "before_ss") {
    title("Northern Cardinal eBird Observations\nBefore subsampling")
  } else {
    title("After subsampling")
  }
}
par(p)
```

```{r}
write.csv(ebird_habitat,'../data/processed/ebird_habitat.csv', row.names=FALSE)
head(ebird_habitat)
write.csv(ebird_ss,'../data/processed/ebird_ss.csv', row.names=FALSE)
head(ebird_ss)

```

## 4.4 Random Forest
```{r}
ebird_ss <- read.csv('../data/processed/ebird_ss.csv')
ebird_split <- ebird_ss %>% 
  # select only the columns to be used in the model
  select(species_observed,
         year, day_of_year,
         time_observations_started, duration_minutes,
         effort_distance_km, number_observers, 
         starts_with("pland_"),
         starts_with("elevation_")) %>% 
  drop_na()
# split 80/20
ebird_split <- ebird_split %>% 
  split(if_else(runif(nrow(.)) <= 0.8, "train", "test"))
map_int(ebird_split, nrow)
```

```{r}
detection_freq <- mean(ebird_split$train$species_observed)
detection_freq
```
```{r}
# ranger requires a factor response to do classification
ebird_split$train$species_observed <- factor(ebird_split$train$species_observed)
head(ebird_split)
write.csv(ebird_split$train, "../data/processed/ebird_split_train.csv", na = "", row.names=FALSE)
write.csv(ebird_split$test, "../data/processed/ebird_split_test.csv", na = "", row.names=FALSE)

# grow random forest
rf <- ranger(formula =  species_observed ~ ., 
             num.trees = 200,
             mtry = 5,
             data = ebird_split$train,
             importance = "impurity",
             probability = TRUE,
             replace = TRUE, 
             sample.fraction = c(detection_freq, detection_freq))

rf
```