---
title: "statistical learning project"
author: "Arjuna Anilkumar, A20446963"
date: "11/12/2020"
output: html_document
---


remotes::install_github("mstrimas/ebppackages")

install.packages(c("dggridR","lubridate","ranger","scam","PresenceAbsence","verification","ebirdst","fields","auk","lubridate","raster","MODIS","exactextractr","viridis","tidyverse","sf","rnaturalearth","dplyr","rgdal","xfun"))
```{r}
# libraries
#library(c("lubridate","ranger","scam","PresenceAbsence","verification","ebirdst","fields","auk","lubridate","raster","MODIS","exactextractr","viridis","tidyverse","sf","rnaturalearth","dplyr","rgdal","xfun"))
```

```{r}
library(sf)
library(rnaturalearth)
library(dplyr)
gpkg_dir <- "../data/raw"
if (!dir.exists(gpkg_dir)) {
  dir.create(gpkg_dir)
}
f_ne <- file.path(gpkg_dir, "gis-data.gpkg")

tmp_dir <- normalizePath("C:\\Users\\arjun\\OneDrive\\MATH 569\\Project\\birding-statistical-learning\\data\\raw\\BCR\\")
tmp_bcr <- normalizePath("C:\\Users\\arjun\\OneDrive\\MATH 569\\Project\\birding-statistical-learning\\data\\raw\\BCR\\bcr_terrestrial_shape.zip")
paste0("https://www.birdscanada.org/download/gislab/bcr_terrestrial_shape.zip") %>% download.file(destfile = tmp_bcr)
unzip(tmp_bcr, exdir = tmp_dir)
bcr <- file.path(tmp_dir, "BCR_Terrestrial_master_International.shp") %>% 
  read_sf() %>% 
  select(bcr_code = BCR, bcr_name = LABEL) %>% 
  filter(bcr_code == 22)

# clean up
list.files(tmp_dir, "bcr", ignore.case = TRUE, full.names = TRUE) %>% 
  unlink()

# political boundaries
# land border with lakes removed
ne_land <- ne_download(scale = 50, category = "cultural",
                       type = "admin_0_countries_lakes",
                       returnclass = "sf") %>%
  filter(CONTINENT == "North America") %>%
  st_set_precision(1e6) %>%
  st_union()
# country lines
# downloaded globally then filtered to north america with st_intersect()
ne_country_lines <- ne_download(scale = 50, category = "cultural",
                                type = "admin_0_boundary_lines_land",
                                returnclass = "sf") %>% 
  st_geometry()
ne_country_lines <- st_intersects(ne_country_lines, ne_land, sparse = FALSE) %>%
  as.logical() %>%
  {ne_country_lines[.]}
# states, north america
ne_state_lines <- ne_download(scale = 50, category = "cultural",
                              type = "admin_1_states_provinces_lines",
                              returnclass = "sf") %>%
  filter(adm0_a3 %in% c("USA")) %>%
  mutate(iso_a2 = recode(adm0_a3, USA = "US")) %>% 
  select(country = adm0_name, country_code = iso_a2)

# output
unlink(f_ne)
write_sf(ne_land, f_ne, "ne_land")
write_sf(ne_country_lines, f_ne, "ne_country_lines")
write_sf(ne_state_lines, f_ne, "ne_state_lines")
write_sf(bcr, f_ne, "bcr")
```



```{r}
library(auk)
library(lubridate)
library(gridExtra)
library(tidyverse)
ebd <- auk_ebd("../data/raw/ebd_US-IL_norcar_relSep-2020.txt")

# filters
ebd_filters <- ebd %>% 
  auk_species("Northern Cardinal") %>% 
  # southeastern coastal plain bcr
  auk_bcr(bcr = 22) %>% 
  # june, use * to get data from any year
  auk_date(date = c("*-06-01", "*-06-30")) %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()
ebd_filters


data_dir <- "C:/Users/arjun/OneDrive/MATH 569/Project/birding-statistical-learning/data/raw"

f_ebd <- file.path(data_dir, "ebd_cardinal_june_bcr22.txt")


if (!file.exists(f_ebd)) {
  auk_filter(ebd_filters, file = f_ebd)
}
```

```{r}
ebd_zf <- read_ebd(f_ebd, unique = TRUE, rollup = TRUE)

time_to_decimal <- function(x) {
  x <- hms(x, quiet = TRUE)
  hour(x) + minute(x) / 60 + second(x) / 3600
}

ebd_zf <- ebd_zf %>% 
  mutate(
    # convert X to NA
    observation_count = if_else(observation_count == "X", 
                                NA_character_, observation_count),
    observation_count = as.integer(observation_count),
    # effort_distance_km to 0 for non-travelling counts
    effort_distance_km = if_else(protocol_type != "Traveling", 
                                 0, effort_distance_km),
    # convert time to decimal hours since midnight
    time_observations_started = time_to_decimal(time_observations_started),
    # split date into year and day of year
    year = year(observation_date),
    day_of_year = yday(observation_date)
  )

ebd_zf_filtered <- ebd_zf %>% 
  filter(
    # effort filters
    duration_minutes <= 5 * 60,
    effort_distance_km <= 5,
    # last 10 years of data
    year >= 2010,
    # 10 or fewer observers
    number_observers <= 10)

ebird <- ebd_zf_filtered %>% 
  select(checklist_id, observer_id, sampling_event_identifier,
         scientific_name,
         observation_count, 
         state_code, locality_id, latitude, longitude,
         protocol_type, all_species_reported,
         observation_date, year, day_of_year,
         time_observations_started, 
         duration_minutes, effort_distance_km,
         number_observers)

ebird <- ebird %>%
  mutate(species_observed = case_when(observation_count >=1 ~ TRUE, TRUE ~ FALSE))

write_csv(ebird, "../data/processed/ebd_cardinal_june_bcr22_zf.csv", na = "")
```

```{r}


# load and project gis data
new_vector = st_read(f_ne)

map_proj <- st_crs(new_vector)
ne_land <- read_sf(f_ne, "ne_land") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
bcr <- read_sf(f_ne, "bcr") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
ne_country_lines <- read_sf(f_ne, "ne_country_lines") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
ne_state_lines <- read_sf(f_ne, "ne_state_lines") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()

# prepare ebird data for mapping
ebird_sf <- ebird %>% 
  # convert to spatial points
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(crs = map_proj) %>% 
  select(species_observed)

# map
par(mar = c(0.25, 0.25, 0.25, 0.25))
  # set up plot area
plot(st_geometry(ebird_sf), col = NA)
  # contextual gis data
plot(ne_land, col = "#dddddd", border = "#888888", lwd = 0.5, add = TRUE)
plot(bcr, col = "#cccccc", border = NA, add = TRUE)
plot(ne_state_lines, col = "#ffffff", lwd = 0.75, add = TRUE)
plot(ne_country_lines, col = "#ffffff", lwd = 1.5, add = TRUE)
  # ebird observations
    # not observed
plot(st_geometry(ebird_sf),
     pch = 19, cex = 0.1, col = alpha("#555555", 0.25),
     add = TRUE)
    # observed
plot(filter(ebird_sf, species_observed) %>% st_geometry(),
     pch = 19, cex = 0.3, col = alpha("#4daf4a", 1),
     add = TRUE)
  # legend
legend("bottomright", bty = "n",
       col = c("#555555", "#4daf4a"),
       legend = c("eBird checklists", "cardinal sightings"),
       pch = 19)
box()
par(new = TRUE, mar = c(0, 0, 3, 0))
title("Cardinal eBird Observations\nJune 2010-2019, BCR 22")
```

# Time of day
```{r}
# summarize data by hourly bins
breaks <- 0:24
labels <- breaks[-length(breaks)] + diff(breaks) / 2
ebird_tod <- ebird %>% 
  mutate(tod_bins = cut(time_observations_started, 
                        breaks = breaks, 
                        labels = labels,
                        include.lowest = TRUE),
         tod_bins = as.numeric(as.character(tod_bins))) %>% 
  group_by(tod_bins) %>% 
  summarise(n_checklists = n(),
            n_detected = sum(species_observed),
            det_freq = mean(species_observed))

# histogram
g_tod_hist <- ggplot(ebird_tod) +
  aes(x = tod_bins, y = n_checklists) +
  geom_segment(aes(xend = tod_bins, y = 0, yend = n_checklists),
               color = "grey50") +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 24, by = 3), limits = c(0, 24)) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Hours since midnight",
       y = "# checklists",
       title = "Distribution of observation start times")

# frequency of detection
g_tod_freq <- ggplot(ebird_tod %>% filter(n_checklists > 100)) +
  aes(x = tod_bins, y = det_freq) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 24, by = 3), limits = c(0, 24)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Hours since midnight",
       y = "% checklists with detections",
       title = "Detection frequency")

# combine
grid.arrange(g_tod_hist, g_tod_freq)
```

# checklist duration

```{r}
# summarize data by 30 minute bins
breaks <- seq(0, 5, by = 0.5)
labels <- breaks[-length(breaks)] + diff(breaks) / 2
ebird_dur <- ebird %>% 
  mutate(dur_bins = cut(duration_minutes / 60, 
                        breaks = breaks, 
                        labels = labels,
                        include.lowest = TRUE),
         dur_bins = as.numeric(as.character(dur_bins))) %>% 
  group_by(dur_bins) %>% 
  summarise(n_checklists = n(),
            n_detected = sum(species_observed),
            det_freq = mean(species_observed))

# histogram
g_dur_hist <- ggplot(ebird_dur) +
  aes(x = dur_bins, y = n_checklists) +
  geom_segment(aes(xend = dur_bins, y = 0, yend = n_checklists),
               color = "grey50") +
  geom_point() +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Checklist duration (hours)",
       y = "# checklists",
       title = "Distribution of checklist durations")

# frequency of detection
g_dur_freq <- ggplot(ebird_dur %>% filter(n_checklists > 100)) +
  aes(x = dur_bins, y = det_freq) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Checklist duration (hours)",
       y = "% checklists with detections",
       title = "Detection frequency")

# combine
grid.arrange(g_dur_hist, g_dur_freq)
```

# Distance travelled

```{r}
# summarize data by 500m bins
breaks <- seq(0, 5, by = 0.5)
labels <- breaks[-length(breaks)] + diff(breaks) / 2
ebird_dist <- ebird %>% 
  mutate(dist_bins = cut(effort_distance_km, 
                         breaks = breaks, 
                         labels = labels,
                         include.lowest = TRUE),
         dist_bins = as.numeric(as.character(dist_bins))) %>% 
  group_by(dist_bins) %>% 
  summarise(n_checklists = n(),
            n_detected = sum(species_observed),
            det_freq = mean(species_observed))

# histogram
g_dist_hist <- ggplot(ebird_dist) +
  aes(x = dist_bins, y = n_checklists) +
  geom_segment(aes(xend = dist_bins, y = 0, yend = n_checklists),
               color = "grey50") +
  geom_point() +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Distance travelled (km)",
       y = "# checklists",
       title = "Distribution of distance travelled")

# frequency of detection
g_dist_freq <- ggplot(ebird_dist %>% filter(n_checklists > 100)) +
  aes(x = dist_bins, y = det_freq) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Distance travelled (km)",
       y = "% checklists with detections",
       title = "Detection frequency")

# combine
grid.arrange(g_dist_hist, g_dist_freq)
```

# No of observers

```{r}
# summarize data
breaks <- 0:10
labels <- 1:10
ebird_obs <- ebird %>% 
  mutate(obs_bins = cut(number_observers, 
                        breaks = breaks, 
                        label = labels,
                        include.lowest = TRUE),
         obs_bins = as.numeric(as.character(obs_bins))) %>% 
  group_by(obs_bins) %>% 
  summarise(n_checklists = n(),
            n_detected = sum(species_observed),
            det_freq = mean(species_observed))

# histogram
g_obs_hist <- ggplot(ebird_obs) +
  aes(x = obs_bins, y = n_checklists) +
  geom_segment(aes(xend = obs_bins, y = 0, yend = n_checklists),
               color = "grey50") +
  geom_point() +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "# observers",
       y = "# checklists",
       title = "Distribution of the number of observers")

# frequency of detection
g_obs_freq <- ggplot(ebird_obs) +
  aes(x = obs_bins, y = det_freq) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "# observers",
       y = "% checklists with detections",
       title = "Detection frequency")

# combine
grid.arrange(g_obs_hist, g_obs_freq)
```


# modis data 



```{r}
library(sf)
library(raster)
library(MODIS)
library(exactextractr)
library(viridis)
library(tidyverse)
# resolve namespace conflicts
select <- dplyr::select
map <- purrr::map
projection <- raster::projection

# bcr 22 boundary
bcr <- read_sf("../data/raw/gis-data.gpkg", "bcr") %>% 
  filter(bcr_code == 22) %>% 
  # project to the native modis projection
  st_transform(crs = paste("+proj=sinu +lon_0=0 +x_0=0 +y_0=0",
                           "+a=6371007.181 +b=6371007.181 +units=m +no_defs"))
# load ebird data
ebird <- read_csv("../data/processed/ebd_cardinal_june_bcr22_zf.csv")
# get list of tiles required to cover this bcr
tiles <- getTile(bcr)
tiles@tile
```
# MODIS setup
```{r}
MODIS:::checkTools("GDAL")
MODIS::MODISoptions(gdalPath = "c:/OSGeo4W64/bin")
```

```{r}
# earliest year of ebird data
begin_year <- format(min(ebird$observation_date), "%Y.01.01")
# end date for ebird data
end_year <- format(max(ebird$observation_date), "%Y.12.31")
# download tiles and combine into a single raster for each year
tifs <- runGdal(product = "MCD12Q1", collection = "006", SDSstring = "01", 
                extent = bcr %>% st_buffer(dist = 10000), 
                begin = begin_year, end = end_year, 
                outDirPath = "../data/raw",job = "modis",
                MODISserverOrder = "LPDAAC")%>% 
  pluck("MCD12Q1.006") %>% 
  unlist()

# rename tifs to have more descriptive names
new_names <- format(as.Date(names(tifs)), "%Y") %>% 
  sprintf("modis_mcd12q1_umd_%s.tif", .) %>% 
  file.path(dirname(tifs), .)
file.rename(tifs, new_names)
```


```{r}
# load the landcover data
landcover <- list.files("../data/raw/modis", "^modis_mcd12q1_umd", 
                        full.names = TRUE) %>% 
  stack()
# label layers with year
landcover <- names(landcover) %>% 
  str_extract("(?<=modis_mcd12q1_umd_)[0-9]{4}") %>% 
  paste0("y", .) %>% 
  setNames(landcover, .)
landcover

max_lc_year <- names(landcover) %>% 
  str_extract("[0-9]{4}") %>% 
  as.integer() %>% 
  max()

neighborhood_radius <- 5 * ceiling(max(res(landcover))) / 2
ebird_buff <- ebird %>% 
  distinct(year = format(observation_date, "%Y"),
           locality_id, latitude, longitude) %>% 
  # for 2019 use 2018 landcover data
  mutate(year_lc = if_else(as.integer(year) > max_lc_year, 
                           as.character(max_lc_year), year),
         year_lc = paste0("y", year_lc)) %>% 
  # convert to spatial features
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  # transform to modis projection
  st_transform(crs = projection(landcover)) %>% 
  # buffer to create neighborhood around each point
  st_buffer(dist = neighborhood_radius) %>% 
  # nest by year
  nest(data = c(year, locality_id, geometry))


calculate_pland <- function(yr, regions, lc) {
  locs <- st_set_geometry(regions, NULL)
  exact_extract(lc[[yr]], regions, progress = FALSE) %>% 
    map(~ count(., landcover = value)) %>% 
    tibble(locs, data = .) %>% 
    unnest(data)
}
# iterate over all years extracting landcover for all checklists in each
lc_extract <- ebird_buff %>% 
  mutate(pland = map2(year_lc, data, calculate_pland, lc = landcover)) %>%
  select(pland) %>% 
  unnest(cols = pland)
```

```{r}
pland <- lc_extract %>% 
  # calculate proporiton
  group_by(locality_id, year) %>% 
  mutate(pland = n / sum(n)) %>% 
  ungroup() %>% 
  select(-n) %>% 
  # remove NAs after tallying so pland is relative to total number of cells
  filter(!is.na(landcover))


# convert names to be more descriptive
lc_names <- tibble(landcover = 0:15,
                   lc_name = c("pland_00_water", 
                               "pland_01_evergreen_needleleaf", 
                               "pland_02_evergreen_broadleaf", 
                               "pland_03_deciduous_needleleaf", 
                               "pland_04_deciduous_broadleaf", 
                               "pland_05_mixed_forest",
                               "pland_06_closed_shrubland", 
                               "pland_07_open_shrubland", 
                               "pland_08_woody_savanna", 
                               "pland_09_savanna", 
                               "pland_10_grassland", 
                               "pland_11_wetland", 
                               "pland_12_cropland", 
                               "pland_13_urban", 
                               "pland_14_mosiac", 
                               "pland_15_barren"))
pland <- pland %>% 
  inner_join(lc_names, by = "landcover") %>% 
  arrange(landcover) %>% 
  select(-landcover)%>% unique()

# tranform to wide format, filling in implicit missing values with 0s%>% 
pland <- pland %>% 
  pivot_wider(names_from = lc_name, 
              values_from = pland, 
              values_fill = list(pland = 0))

# save
write_csv(pland, "../data/raw/modis_pland_location-year.csv")
```

# prediction surface

```{r}
agg_factor <- round(2 * neighborhood_radius / res(landcover))
r <- raster(landcover) %>% 
  aggregate(agg_factor) 
r <- bcr %>% 
  st_transform(crs = projection(r)) %>% 
  rasterize(r, field = 1) %>% 
  # remove any empty cells at edges
  trim()
r <- writeRaster(r, filename = "../data/raw/prediction-surface.tif", overwrite = TRUE)


# get cell centers and create neighborhoods
r_centers <- rasterToPoints(r, spatial = TRUE) %>% 
  st_as_sf() %>% 
  transmute(id = row_number())
r_cells <- st_buffer(r_centers, dist = neighborhood_radius)

# extract landcover values within neighborhoods, only needed most recent year
lc_extract_pred <- landcover[[paste0("y", max_lc_year)]] %>% 
  exact_extract(r_cells, progress = FALSE) %>% 
  map(~ count(., landcover = value)) %>% 
  tibble(id = r_cells$id, data = .) %>% 
  unnest(data)

# calculate the percent for each landcover class
pland_pred <- lc_extract_pred %>% 
  count(id, landcover) %>% 
  group_by(id) %>% 
  mutate(pland = n / sum(n)) %>% 
  ungroup() %>% 
  select(-n) %>% 
  # remove NAs after tallying so pland is relative to total number of cells
  filter(!is.na(landcover))

# convert names to be more descriptive
pland_pred <- pland_pred %>% 
  inner_join(lc_names, by = "landcover") %>% 
  arrange(landcover) %>% 
  select(-landcover)

# tranform to wide format, filling in implicit missing values with 0s
pland_pred <- pland_pred %>% 
  pivot_wider(names_from = lc_name, 
              values_from = pland, 
              values_fill = list(pland = 0)) %>% 
  mutate(year = max_lc_year) %>% 
  select(id, year, everything())

# join in coordinates
pland_coords <- st_transform(r_centers, crs = 4326) %>% 
  st_coordinates() %>% 
  as.data.frame() %>% 
  cbind(id = r_centers$id, .) %>% 
  rename(longitude = X, latitude = Y) %>% 
  inner_join(pland_pred, by = "id")

# plot 
forest_cover <- pland_coords %>% 
  # convert to spatial features
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(crs = projection(r)) %>% 
  # rasterize points
  rasterize(r, field = "pland_04_deciduous_broadleaf") %>% 
  # project to albers equal-area for mapping
  projectRaster(crs = st_crs(new_vector)$proj4string, method = "ngb") %>% 
  # trim off empty edges of raster
  trim()

# make a map
par(mar = c(0.25, 0.25, 2, 0.25))
t <- str_glue("Proportion of Deciduous Broadleaf Forest\n",
              "{max_lc_year} MODIS Landcover")
plot(forest_cover, axes = FALSE, box = FALSE, col = viridis(10), main = t)
```

# Elevation

```{r}
elev <- raster("../data/raw/elevation_1KMmd_GMTEDmd.tif")
# crop, buffer bcr by 10 km to provide a little wiggly room
elev <- bcr %>% 
  st_buffer(dist = 10000) %>% 
  st_transform(crs = projection(elev)) %>% 
  crop(elev, .) %>% 
  projectRaster(crs = projection(landcover))

# buffer each checklist location
ebird_buff_noyear <- ebird %>% 
  distinct(locality_id, latitude, longitude) %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(crs = projection(elev)) %>% 
  st_buffer(dist = neighborhood_radius)

# extract elevation values and calculate median and sd
locs <- st_set_geometry(ebird_buff_noyear, NULL) %>% 
    mutate(id = row_number())
elev_checklists <- exact_extract(elev, ebird_buff_noyear, progress = FALSE) %>% 
  map_dfr(~ tibble(elevation_median = mean(.$value, na.rm = TRUE),
                   elevation_sd = sd(.$value, na.rm = TRUE))) %>% 
  # join to lookup table to get locality_id
  bind_cols(locs, .)

# extract and calculate median and sd
elev_pred <- exact_extract(elev, r_cells, progress = FALSE) %>% 
  map_dfr(~ tibble(elevation_median = mean(.$value, na.rm = TRUE),
                   elevation_sd = sd(.$value, na.rm = TRUE))) %>% 
  # join to lookup table to get locality_id
  bind_cols(st_drop_geometry(r_cells), .)

# checklist covariates
pland_elev_checklist <- inner_join(pland, elev_checklists, by = "locality_id")
write_csv(pland_elev_checklist, "../data/processed/pland-elev_location-year.csv")

# prediction surface covariates
pland_elev_pred <- inner_join(pland_coords, elev_pred, by = "id")
write_csv(pland_elev_pred, "../data/processed/pland-elev_prediction-surface.csv")
```




```{r}
library(sf)
library(raster)
library(lubridate)
library(ranger)
library(scam)
library(PresenceAbsence)
library(verification)
library(ebirdst)
library(fields)
library(gridExtra)
library(tidyverse)

select <- dplyr::select
map <- purrr::map
projection <- raster::projection

# set random number seed to insure fully repeatable results
set.seed(1)

# setup output directory for saved results
if (!dir.exists("output")) {
  dir.create("output")
}

# ebird data
ebird <- read_csv("../data/processed/ebd_cardinal_june_bcr22_zf.csv") %>% 
  # year required to join to habitat data
  mutate(year = year(observation_date))

# modis habitat covariates
habitat <- read_csv("../data/processed/pland-elev_location-year.csv") %>% 
  mutate(year = as.integer(year))

# combine ebird and habitat data
ebird_habitat <- inner_join(ebird, habitat, by = c("locality_id", "year"))

# prediction surface
pred_surface <- read_csv("../data/processed/pland-elev_prediction-surface.csv")
# latest year of landcover data
max_lc_year <- max(pred_surface$year)
r <- raster("../data/raw/prediction-surface.tif")

# load gis data for making maps
map_proj <- st_crs(new_vector)
ne_land <- read_sf("../data/raw/gis-data.gpkg", "ne_land") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
bcr <- read_sf("../data/raw/gis-data.gpkg", "bcr") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
ne_country_lines <- read_sf("../data/raw/gis-data.gpkg", "ne_country_lines") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
ne_state_lines <- read_sf("../data/raw/gis-data.gpkg", "ne_state_lines") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()

ebird_split_train <- read.table("../data/processed/ebird_split_train.csv", sep = ",", header = T)
ebird_split_train <- ebird_split_train[,-1]
ebird_split_test <- read.table("../data/processed/ebird_split_test.csv", sep = ",", header = T)
ebird_split_test <- ebird_split_test[,-1]
head(ebird_split_train)
head(ebird_split_test)

detection_freq <- mean(ebird_split_train$species_observed)
# grow random forest
rf <- ranger(formula =  species_observed ~ ., 
             data = ebird_split_train,
             importance = "impurity",
             probability = TRUE,
             replace = TRUE)
rf
```

# Calibration

```{r}
# make predictions on training data
occ_pred <- rf$predictions[, 2]
# convert the observered response back to a numeric value from factor
occ_obs <- ebird_split_train$species_observed %>% 
  as.logical() %>% 
  as.integer()
rf_pred_train <- tibble(obs = occ_obs, pred = occ_pred) %>% 
  drop_na()

# fit calibration model
calibration_model <- scam(obs ~ s(pred, k = 5, bs = "mpi"), 
                          gamma = 1.4,
                          data = rf_pred_train)

# calculate the average observed encounter rates for different 
# categories of estimated encounter rates 

average_encounter <- rf_pred_train %>%
  mutate(pred_cat = cut(rf_pred_train$pred, breaks = seq(0, 1, by=0.02))) %>%
  group_by(pred_cat) %>%
  summarise(pred = mean(pred), obs = mean(obs), checklist_count = n()) %>%
  ungroup()

# plot
cal_pred <- tibble(pred = seq(0, 1, length.out = 100))
cal_pred <- predict(calibration_model, cal_pred, type = "response") %>% 
  bind_cols(cal_pred, calibrated = .)
ggplot(cal_pred) +
  aes(x = pred, y = calibrated) +
  geom_line() +
  geom_point(data = average_encounter, 
             aes(x = pred, y = obs, size = sqrt(checklist_count)),
             show.legend = FALSE, shape = 1) +
  labs(x = "Estimated encounter rate",
       y = "Observed encounter rate",
       title = "Calibration model")
```

# Assessment

```{r}
# predict on test data using calibrated model
p_fitted <- predict(rf, data = ebird_split_test, type = "response")
# extract probability of detection
p_fitted <- p_fitted$predictions[, 2]
# calibrate
p_calibrated <- predict(calibration_model, 
                        newdata = tibble(pred = p_fitted), 
                        type = "response")
rf_pred_test <- data.frame(id = seq_along(p_calibrated),
                           # actual detection/non-detection
                           obs = ebird_split_test$species_observed,
                           # uncalibrated prediction
                           fit = p_fitted,
                           # calibrated prediction
                           cal = p_calibrated) %>%
  # constrain probabilities to 0-1
  mutate(cal = pmin(pmax(cal, 0), 1)) %>% 
  drop_na()

# mean squared error (mse)
mse_fit <- mean((rf_pred_test$obs - rf_pred_test$fit)^2, na.rm = TRUE)
mse_cal <- mean((rf_pred_test$obs - rf_pred_test$cal)^2, na.rm = TRUE)

# pick threshold to maximize kappa
opt_thresh <- optimal.thresholds(rf_pred_test, opt.methods = "MaxKappa")

# calculate accuracy metrics: auc, kappa, sensitivity, specificity,
metrics_fit <- rf_pred_test %>% 
  select(id, obs, fit) %>% 
  presence.absence.accuracy(threshold = opt_thresh$fit, 
                            na.rm = TRUE, 
                            st.dev = FALSE)
metrics_cal <- rf_pred_test %>% 
  select(id, obs, cal) %>% 
  presence.absence.accuracy(threshold = opt_thresh$cal, 
                            na.rm = TRUE, 
                            st.dev = FALSE)

rf_assessment <- tibble(
  model = c("RF", "Calibrated RF"),
  mse = c(mse_fit, mse_cal),
  sensitivity = c(metrics_fit$sensitivity, metrics_cal$sensitivity),
  specificity = c(metrics_fit$specificity, metrics_cal$specificity),
  auc = c(metrics_fit$AUC, metrics_cal$AUC),
  kappa = c(metrics_fit$Kappa, metrics_cal$Kappa)
)
knitr::kable(rf_assessment, digits = 3)
```



```{r}
pi <- enframe(rf$variable.importance, "predictor", "importance")
# plot
ggplot(pi) + 
  aes(x = fct_reorder(predictor, importance), y = importance) +
  geom_col() +
  geom_hline(yintercept = 0, size = 2, colour = "#555555") +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip() +
  labs(x = NULL, 
       y = "Predictor Importance (Gini Index)") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        panel.grid.major.x = element_line(colour = "#cccccc", size = 0.5))

# top 9 predictors other than date
top_pred <- pi %>% 
  filter(!predictor %in% c("year", "day_of_year")) %>% 
  top_n(n = 9, wt = importance) %>% 
  arrange(desc(importance))
```


# Partial dependence

```{r}
# function to calculate partial dependence for a single predictor
calculate_pd <- function(predictor, model, data, 
                         x_res = 25, n = 1000) {
  # create prediction grid
  rng <- range(data[[predictor]], na.rm = TRUE)
  x_grid <- seq(rng[1], rng[2], length.out = x_res)
  grid <- data.frame(covariate = predictor, x = x_grid, 
                     stringsAsFactors = FALSE)
  names(grid) <- c("covariate", predictor)
  
  # subsample training data
  n <- min(n, nrow(data))
  s <- sample(seq.int(nrow(data)), size = n, replace = FALSE)
  data <- data[s, ]
  
  # drop focal predictor from data
  data <- data[names(data) != predictor]
  grid <- merge(grid, data, all = TRUE)
  
  # predict
  p <- predict(model, data = grid)
  
  # summarize
  pd <- grid[, c("covariate", predictor)]
  names(pd) <- c("covariate", "x")
  pd$pred <- p$predictions[, 2]
  pd <- dplyr::group_by(pd, covariate, x) %>% 
    dplyr::summarise(pred = mean(pred, na.rm = TRUE)) %>% 
    dplyr::ungroup()
  
  return(pd)
}

# calculate partial dependence for each predictor
# map is used to iteratively apply calculate_pd to each predictor
pd <- top_pred %>% 
  mutate(pd = map(predictor, calculate_pd, model = rf, 
                  data = ebird_split_train),
         pd = map(pd, ~ .[, c(2, 3)]),
         pd = map(pd, set_names, nm = c("value",  "encounter_rate"))) %>% 
  unnest(cols = pd)

# calibrate predictions
pd$encounter_rate <- predict(calibration_model, 
                             newdata = tibble(pred = pd$encounter_rate), 
                             type = "response") %>% 
  as.numeric()
  
# plot
ggplot(pd) +
  aes(x = value, y = encounter_rate) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ as_factor(predictor), nrow = 3, scales = "free") +
  labs(x = NULL, y = "Encounter Rate") +
  theme_minimal() +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.line = element_line(color = "grey60"),
        axis.ticks  = element_line(color = "grey60"))
```

```{r}
# find peak time of day from partial dependence
pd_time <- calculate_pd("time_observations_started",
                        model = rf, 
                        data = ebird_split_train,
                        # make estimates at 30 minute intervals
                        # using a subset of the training dataset
                        x_res = 2 * 24, n = 1000) %>% 
  transmute(time_observations_started = x, encounter_rate = pred)

# histogram
g_hist <- ggplot(ebird_split_train) +
  aes(x = time_observations_started) +
  geom_histogram(binwidth = 1, center = 0.5, color = "grey30",
                 fill = "grey50") +
  scale_x_continuous(breaks = seq(0, 24, by = 3)) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Hours since midnight",
       y = "# checklists",
       title = "Distribution of observation start times")

# gam
g_pd <- ggplot(pd_time) +
  aes(x = time_observations_started, y = encounter_rate) +
  geom_line() +
  scale_x_continuous(breaks = seq(0, 24, by = 3)) +
  labs(x = "Hours since midnight",
       y = "Probability of reporting",
       title = "Observation start time partial dependence")

# combine
grid.arrange(g_hist, g_pd)
```

```{r}
# hours with at least 1% of checklists
search_hours <- ebird_split_train %>% 
  mutate(hour = floor(time_observations_started)) %>%
  count(hour) %>% 
  mutate(pct = n / sum(n)) %>% 
  filter(pct >= 0.01)

# constrained peak time
t_peak <- pd_time %>% 
  filter(floor(time_observations_started) %in% search_hours$hour) %>% 
  top_n(1, wt = desc(time_observations_started)) %>% 
  pull(time_observations_started)
t_peak
#> [1] 5.07
```

```{r}
# add effort covariates to prediction 
pred_surface_eff <- pred_surface %>% 
  mutate(observation_date = ymd(str_glue("{max_lc_year}-06-15")),
         year = year(observation_date),
         day_of_year = yday(observation_date),
         time_observations_started = t_peak,
         duration_minutes = 60,
         effort_distance_km = 1,
         number_observers = 1)

# predict
pred_rf <- predict(rf, data = pred_surface_eff, type = "response")
pred_rf <- pred_rf$predictions[, 2]
# apply calibration models
pred_rf_cal <- predict(calibration_model, 
                       data.frame(pred = pred_rf), 
                       type = "response")
# add to prediction surface
pred_er <- bind_cols(pred_surface_eff, encounter_rate = pred_rf_cal) %>% 
  select(latitude, longitude, encounter_rate) %>% 
  mutate(encounter_rate = pmin(pmax(encounter_rate, 0), 1))


r_pred <- pred_er %>% 
  # convert to spatial features
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(crs = projection(r)) %>% 
  # rasterize
  rasterize(r)
r_pred <- r_pred[[-1]]

# save the raster
tif_dir <- "output"
if (!dir.exists(tif_dir)) {
  dir.create(tif_dir)
}
writeRaster(r_pred, file.path(tif_dir, "rf-model_encounter-rate_cardinal.tif"), 
            overwrite = TRUE)
```

```{r}
# project predictions
r_pred_proj <- projectRaster(r_pred, crs = map_proj$proj4string, method = "ngb")

par(mar = c(3.5, 0.25, 0.25, 0.25))
# set up plot area
plot(bcr, col = NA, border = NA)
plot(ne_land, col = "#dddddd", border = "#888888", lwd = 0.5, add = TRUE)

# encounter rate
r_max <- ceiling(10 * cellStats(r_pred_proj, max)) / 10
brks <- seq(0, r_max, by = 0.025)
lbl_brks <- seq(0, r_max, by = 0.1)
# ebird status and trends color palette
pal <- abundance_palette(length(brks) - 1)
plot(r_pred_proj, 
     col = pal, breaks = brks, 
     maxpixels = ncell(r_pred_proj),
     legend = FALSE, add = TRUE)

# borders
plot(bcr, border = "#000000", col = NA, lwd = 1, add = TRUE)
plot(ne_state_lines, col = "#ffffff", lwd = 0.75, add = TRUE)
plot(ne_country_lines, col = "#ffffff", lwd = 1.5, add = TRUE)
box()

# legend
par(new = TRUE, mar = c(0, 0, 0, 0))
title <- "Cardinal Encounter Rate"
image.plot(zlim = range(brks), legend.only = TRUE, 
           col = pal, breaks = brks,
           smallplot = c(0.25, 0.75, 0.06, 0.09),
           horizontal = TRUE,
           axis.args = list(at = lbl_brks, labels = lbl_brks,
                            fg = "black", col.axis = "black",
                            cex.axis = 0.75, lwd.ticks = 0.5,
                            padj = -1.5),
           legend.args = list(text = title,
                              side = 3, col = "black",
                              cex = 1, line = 0))
```


